The Effect of a Second AIDS Defining Illness on Survival in the MACS

### 2015-05-04 paper.Rmd
Add lots of confounders to data for model. Do adjusted PH

# read in visit level data
visits <- read_csv("data/hivvert-data04.csv",
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # put visits in correct 1,2, 3,... form
    mutate(visit=VISIT / 10,
           visit.date=mdy(VDATE),
           haart.era.visit=year(visit.date) <= 1994) %>%

@@

    # keep MACS only
    filter(COHORT==1,
           # keep people with first ADI only
           frst.aids.diag != 0,
           # restrict to haart era visits right now
           haart.era.visit == TRUE) %>%
    select(id=STUDYID, center=CENTER, visit:dead.at.censor)
```

```{r fig1}
# therapy <- visits %>%
#     group_by(year(visit.date)) %>%
#     summarise(therapy=list(table(THRPY))) %>%
#     unnest(therapy)

data.plot <- visits %>%
    group_by(id) %>%
    summarise(sec.aids=min(sec.aids.diag)) %>%
    mutate(sec.aids=plyr::revalue(as.character(sec.aids),
                                  c("0"="No Aids",
                                    "1"="PCP",
                                    "2"="KS",
                                    "3"="NHL",
                                    "4"="Neuro",
                                    "5"="Bacteria OI",
                                    "6"="Fungal OI",
                                    "7"="Protozoa",
                                    "8"="CMV/HSV",
                                    "9"="Wasting",
                                    "10"="Recurrent PNA",
                                    "11"="Cervical Ca"),
                                  warn_missing=FALSE))
ggplot(data.plot, aes(x=sec.aids)) + 
    geom_histogram() +
    xlab("Type of Secondary Aids Diagnosis") +
    ylab("Count") +
    ggtitle("Figure 1") +
    scale_x_discrete(limits=names(sort(-table(data.plot$sec.aids)))) +
    theme(axis.text.x = element_text(angle=-90))
```

@@

@@

```{r strat-surv}
# strat.model <- coxph(Surv(first.aids.date, stop, dead.at.censor) ~ 
#                      had.sec.aids + strata(haart.era),
#                      data=data)
```

```{r propensity-score}
# fit logit model (and nonparam?) to calculate prop score
```

```{r matching}
```

>>>>

# read in visit level data
visits <- read_csv("data/hivvert-data04.csv",
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # keep MACS only
    filter(COHORT==1) %>%
    select(STUDYID) %>%
    distinct(STUDYID)

covariates <- covariates %>%
    inner_join(visits)
```

```{r fig1}
visits <- read_csv("data/hivvert-data04.csv",
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # put visits in correct 1,2, 3,... form
    mutate(visit=VISIT / 10,
           visit.date=mdy(VDATE),
           haart.era.visit=year(visit.date) <= 1994) %>%

@@

    # keep MACS only
    filter(COHORT==1,
           # keep people with first ADI only
           frst.aids.diag != 0,
           # restrict to haart era visits right now
           haart.era.visit == TRUE) %>%
    select(id=STUDYID, center=CENTER, visit:dead.at.censor)

# only for pre haart!
# need to do again for all data
data.plot <- visits %>%
    group_by(id) %>%
    summarise(sec.aids=min(sec.aids.diag)) %>%
    mutate(sec.aids=plyr::revalue(as.character(sec.aids),
                                  c("0"="No Aids",
                                    "1"="PCP",
                                    "2"="KS",
                                    "3"="NHL",
                                    "4"="Neuro",
                                    "5"="Bacteria OI",
                                    "6"="Fungal OI",
                                    "7"="Protozoa",
                                    "8"="CMV/HSV",
                                    "9"="Wasting",
                                    "10"="Recurrent PNA",
                                    "11"="Cervical Ca"),
                                  warn_missing=FALSE))
ggplot(data.plot, aes(x=sec.aids)) + 
    geom_histogram() +
    xlab("Type of Secondary Aids Diagnosis") +
    ylab("Count") +
    ggtitle("Figure 1") +
    scale_x_discrete(limits=names(sort(-table(data.plot$sec.aids)))) +
    theme(axis.text.x = element_text(angle=-90))
```

@@

knitr::kable(coef(summary(naive.model)),
             caption="Naive Model (All Data)")
```

```{r all-data}
# read in horiz data with lots of covariates
covariates <- read_csv("data/macswihs_horiz2013_copy-1.csv") %>%
           # date of death
    mutate(death.date=as.Date("1960-01-01") + FRSTDTHD,
           # date of aids diagnosis
           aids.date=as.Date("1960-01-01") + FRSTAIDD,
           # date of secondary aids diagnosis
           sec.aids.date=as.Date("1960-01-01") + SECAIDD,
           # what was the first aids diagnosis
           frst.aids.diag=AIDSDIAG,
           # what was the second aids diagnosis
           sec.aids.diag=SECAIDSDIAG,
           # did the person have a sec aids diag
           had.sec=year(sec.aids.date) < 2100,
           # did the person die
           died=year(death.date) < 2100,
           dob=as.Date("1960-01-01") + DOB,
           age.at.aids=difftime(aids.date, dob, unit="weeks") / 52.25,
           age.at.aids=as.numeric(age.at.aids),
           exit=death.date,
           censor.date=as.Date("1970-01-01") + as.integer(exit),
           last.alive=as.Date("1960-01-01") + LASTALID,
           censor.date=ifelse(last.alive < censor.date, last.alive, 
                              censor.date),
           censor.date=as.Date("1970-01-01")+censor.date) %>%
    select(STUDYID, 
           aids.date,
           sec.aids.date,
           frst.aids.diag, 
           censor.date, 
           sec.aids.diag,
           death.date,
           race=RACE,
           age.at.aids) %>%
    filter(year(aids.date) < 2100) %>%
    mutate(haart.era=(year(aids.date) > 1993))

# what are the important confoudners that we can include for the PH model?
# what about propensity score?

# what about therapy?
visits <- read_csv("data/hivvert-data04.csv",
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # keep MACS only
    filter(COHORT == 1) %>% 
    group_by(STUDYID) %>%
    summarise(drug.use=sum(POTUSE, CRACKUSE, COCUSE, HERUSE, AMPHUSE, 
                           POPUSE, na.rm=TRUE) > 1,
              depression=mean(CESD, na.rm=TRUE) < ifelse(drug.use, 
                                                         16, 23))

covariates <- covariates %>%
    mutate(aids.year=year(aids.date),
           aids.month=month(aids.date))

# cd4 closest to primary AIDS diagnosis
cd4 <- read_csv("data/hivvert-data04.csv",
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    filter(COHORT == 1, !is.na(CD4N)) %>%
    select(STUDYID, cd4=CD4N, VDATE) %>%
    mutate(VDATE=mdy(VDATE),
           VDATE=as.Date(VDATE)) %>%
    inner_join(covariates, by=c("STUDYID"="STUDYID")) %>%
    select(STUDYID, cd4, VDATE, aids.date) %>%
    mutate(visit.aids.diff=difftime(VDATE, aids.date)) %>%
    group_by(STUDYID) %>%
    filter(visit.aids.diff == min(abs(visit.aids.diff))) %>%
    select(STUDYID, cd4)

# join in cd4 at first AIDS diagnosis
summ.visits <- visits %>%
    left_join(cd4)

# add data together
covariates <- covariates %>%
    inner_join(summ.visits) %>%
    mutate(race=ifelse(race == 1, "white", "non-white"))
```

We also created a naive model that includes **all** MACS data,
but does not adjust for any confounders, or the pre / post HAART era.
```{r all-naive-model}
data <- covariates %>%
    rowwise() %>%
    mutate(start=aids.date,
           stop=censor.date,
           sec.aids=sec.aids.date,
           status=(death.date < as.Date("2100-01-01"))) %>%
    mutate_each(funs(as.numeric), start, stop, sec.aids) %>%
    # there are a few obs coming up where secondary and primary are 
    # diagnosed at the same time.
    # let's remove those.
    filter(stop > start,
           !is.na(sec.aids))

@@

Next, we adjusted for several confounders, such as Race, Age at baseline, CD4 at baseline, year of AIDS diagnosis, and drug use.

```{r all-adjusted}
adjusted.model <- coxph(Surv(start, stop, status) ~ sec.aids +
                        factor(race) + age.at.aids + drug.use +
                        aids.year + depression + cd4,
                        data=time.varying)
```

```{r propensity-score, eval=FALSE}
prop.score <- glm(sec.aids ~ confounder1 + confounder2,
                  family=binomial,
                  data=time.varying)
# prop.score <- glm(sec.aids ~ status + haart.era,
#                   family=binomial,
#                   data=time.varying)

time.varying <- time.varying %>%
    mutate(propensity=predict(prop.score, type="response"))
```

```{r matching, eval=FALSE}
library(MatchIt)
```

```{r sophisticated-hazards}
```

```{r graphs}
```
